---
title: "**TI-84 Summary Statistics and Graphs**"
author: "**Arnab Aich**"
output: 
  html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cerulean
---
<link rel="stylesheet" type="text/css" href="../styles.css">

## Introduction

In this document, we will guide you through calculating various summary statistics and generating graphs using the TI-84 calculator. This includes calculating the **mean**, **median**, **standard deviation**, **range**, **IQR**, and **Z-scores**, as well as creating **bar plots**, **pie charts**, **histograms**, and **boxplots**.

We will use the following example dataset for all calculations:

- Values: 2, 5, 7, 9, 12, 15, 18, 21, 25

## <span style="color:#7C0A02;">Calculating Summary Measures on TI-84:</span>

### Steps:
1. **Enter Data:**
   <div class="calc-input">
   Press <code>STAT</code>, then <code>1:Edit</code> to access the list editor.<br>
   Enter the dataset values into List 1 (L1).
   </div>

2. **Calculate Summary Statistics:**
   <div class="calc-input">
   After entering data, press <code>STAT</code>, then scroll right to <code>CALC</code>.<br>
   Select <code>1:1-Var Stats</code> and press <code>ENTER</code>.<br>
   Ensure that <code>L1</code> is selected and press <code>ENTER</code> again.
   </div>

The calculator will display the following values:

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(magrittr)

data <- data.frame(
  Statistic = c("Mean", "Sum of x", "Sample SD", "Population SD",
                "Minimum", "First Quartile", "Median", "Third Quartile", "Maximum"),
  Formula = c("\\( \\bar{x} \\)", "\\( \\sigma x \\)", "\\(  s_x \\)", "\\( \\sigma_x \\)",
              "\\( minX \\)", "\\( Q_1 \\)", "\\( Q_2 \\)", "\\( Q_3 \\)", "\\( \\text{maxX} \\)")
)

kable(data, format = "html", align = "c", escape = FALSE, col.names = c("Statistic", "Formula")) %>%
kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#7C0A02") %>%
  row_spec(1:nrow(data), background = "lightblue")



```
 **Calculated Values from datasets:**

```{r echo=FALSE, message=FALSE, warning=FALSE}

data <- data.frame(
  Statistic = c("Mean", "Median", "SD( $s_x$ )", "IQR (\\( Q_3 - Q_1 \\))"),
  Value = c(12.67, 12, 7.43, 8)
)

kable(data, format = "html", align = "c", escape = FALSE, col.names = c("Statistic", "Value")) %>%
kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#7C0A02") %>%
  row_spec(1:nrow(data), background = "lightblue")

```


## <span style="color:#7C0A02;">Calculating Z-Scores on TI-84:</span>

To calculate the Z-scores for specific values, follow these steps:

1. **Find Mean and Standard Deviation:**
   - Use the steps above to calculate the mean and standard deviation.

2. **Use the Formula:**
   \[
   Z = \frac{X - \mu}{\sigma}
   \]
   Where:
   - \\( X \\) is the data point,
   - \\( \mu \\) is the mean,
   - \\( \sigma \\) is the standard deviation.

3. **Example for Dataset:**
   - For the value 18, the Z-score is calculated as follows:
   \[
   Z = \frac{18 - 12.67}{7.43} \approx 0.72
   \]

## <span style="color:#7C0A02;">Creating Graphs on TI-84:</span>

### 1. **Bar Plot:**
   <div class="calc-input">
   Enter the data into <code>L1</code>.<br>
   Press <code>2nd</code>, then <code>Y=</code> to access the STAT PLOT menu.<br>
   Turn on Plot1 and select the **Histogram** type.<br>
   Choose <code>L1</code> as the data source and press <code>GRAPH</code>.
   </div>

### 2. **Histogram:**
   <div class="calc-input">
   Follow the steps for the bar plot but adjust the <code>Xmin</code>, <code>Xmax</code>, and bin width under <code>WINDOW</code> settings to fit your data range.
   </div>

### 3. **Boxplot:**
   <div class="calc-input">
   Press <code>2nd</code>, then <code>Y=</code> to access the STAT PLOT menu.<br>
   Turn on Plot1 and select the **Boxplot** option.<br>
   Press <code>GRAPH</code> to display the boxplot.
   </div>

---
