---
title: "Interactive Statistical Concepts"
author: "Arnab Aich"
output:
  html_document:
    self_contained: true
    runtime: shiny
    embed-resources: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, input = FALSE, output = TRUE)
# List of libraries to load
libraries <- c("shiny", "plotly", "gtsummary", "gt", "tibble", "shinyjs", "shinyalert", "shinyWidgets", 
               "ggplot2", "datasets", "DT", "BSDA", "knitr", "kableExtra", "statmod")

# Load each library
lapply(libraries, library, character.only = TRUE)


```



<style>
  body {
    font-family: 'Georgia', serif;
    line-height: 1.6;
  }
  h1, h2, h3 {
    font-family: 'Verdana', sans-serif;
  }
  p {
    font-size: 18px;
    color: #212121;
  }
  ul {
    font-family: 'Arial', sans-serif;
    font-size: 16px;
    color: #616161;
  }
  a {
    color: #FF5722;
  }
  .tabset li.active {
    background-color: #4CAF50;
    color: white;
  }
  .app-container {
    margin-top: 20px;
    padding: 20px;
    border: 1px solid #ddd;
    border-radius: 10px;
    background-color: #f9f9f9;
  }
</style>

<p style="color: #2E86C1; font-size: 18px;">
Welcome! This document will guide you through some essential ideas in statistics using interactive and engaging tools. Statistics isn't just about numbers—it's about understanding patterns, making predictions, and drawing meaningful conclusions from data. With these tools, you'll get hands-on experience that makes learning statistics intuitive and fun.
</p>

<p style="color: #2E4053; font-size: 18px;">
These tools are designed to be easy to use and informative. Whether you're new to statistics or just looking to refresh your knowledge, you'll learn by doing—exploring data, seeing how it behaves, and discovering how to draw conclusions. Let's dive in and see what we can learn together!
</p>

## {.tabset .tabset-fade}

### **Preliminaries: Understanding Sampling**

#### **What's This About?**
<p>
Imagine you're in a forest and want to know the average height of the trees. Measuring every single tree would take forever! So, instead, you measure a few trees (this is your "sample") and use that information to guess the average height of all the trees in the forest. This tool lets you see how changing the size of your sample affects your guess. Try it out—see how bigger samples tend to give you better estimates!
</p>

#### **Why It Matters**
<p>
Sampling isn't just something we do with trees. We use sampling to estimate all sorts of things—from how much people earn in a country to how likely voters are to support a candidate in an election. The key is understanding how accurate our estimates are, and this tool helps you get a feel for that by letting you play with different sample sizes.
</p>

#### **How to Use the Tool**
<ul>
  <li><b>Adjust the Sample Size</b>: Use the slider to change how many trees you're measuring. Watch how the average height changes with different sample sizes.</li>
  <li><b>Compare with the True Average</b>: The tool shows you both your estimate and the actual average height of all the trees. See how close you can get!</li>
</ul>

<div class="app-container">
```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
app <- source("R/Preliminaries.R")$value
shinyApp(ui = app$ui, server = app$server)
```
</div>

### **Summary Statistics: The Basics**

#### **What's This About?**
<p>
When you have a lot of data, it can be overwhelming. That's where summary statistics come in—they help you sum up all that data into a few key numbers. For example, instead of listing everyone's scores on a test, you could just tell me the average score, the highest score, and the lowest score. This tool lets you explore these summary statistics and see what they tell us about the data.
</p>

#### **Why It Matters**
<p>
Summary statistics are used everywhere—from healthcare to business. They help us quickly understand and compare data. In this tool, you'll see how different summary statistics, like the mean (average) and standard deviation (spread of data), give us different insights into the same dataset.
</p>

#### **How to Use the Tool**
<ul>
  <li><b>Choose a Dataset</b>: Pick a dataset to analyze.</li>
  <li><b>Explore the Statistics</b>: The tool calculates important numbers like the average and standard deviation for you.</li>
  <li><b>Visualize the Data</b>: Use the plots to see the distribution of the data—this makes it easier to understand what's going on!</li>
</ul>

<div class="app-container">
```{r Summary Statistics}
app <- source("R/Summary_Stat.R")$value
shinyApp(ui = app$ui, server = app$server)
```
</div>

### **Distributions: Exploring Possibilities**

#### **What's This About?**
<p>
A distribution shows us all the possible outcomes of something and how likely each outcome is. Think about rolling a die—each number from 1 to 6 has an equal chance of coming up, and a distribution can show us that. This tool lets you explore different types of distributions and see how they change when you tweak the settings.
</p>

#### **Why It Matters**
<p>
Understanding distributions is super important because they help us predict outcomes and make decisions. For example, in finance, distributions might be used to model stock returns. This tool helps you see how distributions work by letting you adjust things like the mean (center) and standard deviation (spread) and see how the shape of the distribution changes.
</p>

#### **How to Use the Tool**
<ul>
  <li><b>Pick a Distribution</b>: Choose from Normal, Chi-squared, t, or F distributions.</li>
  <li><b>Adjust the Parameters</b>: Change the mean, standard deviation, and other settings to see how the distribution changes.</li>
  <li><b>See the Probabilities</b>: The tool calculates probabilities for you and shows them visually on the distribution plot.</li>
</ul>

<div class="app-container">
```{r Distributions}
app <- source("R/Distributions.R")$value
shinyApp(ui = app$ui, server = app$server)
```
</div>

## **Conclusion**

<p style="color: #2E86C1; font-size: 18px;">
We hope these interactive tools have helped you see how statistics works in a fun and engaging way. Remember, statistics is all about making sense of data, and the more you practice, the better you'll get at it. Keep exploring, keep asking questions, and keep learning—there's a whole world of data out there waiting for you to discover!
</p>

<p style="color: #2E4053; font-size: 18px;">
Whether you're looking to understand data in your studies, your work, or just in everyday life, the concepts you've learned here are the building blocks. So go ahead, dive deeper, and see what else you can discover with the power of statistics!
</p>

<footer style="margin-top: 50px; text-align: center;">
  <p style="font-size: 18px;">Author: Arnab Aich</p>
  <p style="font-size: 16px;">Thanks for exploring statistics with us!</p>
</footer>
